---
title: "Bounds testing when the controls carry trends"
author: "Dr Merwan Roudane"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Bounds testing when the controls carry trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.4, fig.height = 4.0, fig.align = "center",
  dpi = 150
)
options(digits = 4)
library(ardldml)
```

# What this package is for

`ardldml` implements **DML-Bounds**, the procedure of Villena (2026), for
testing whether two series share a long-run relationship *given* a control set
that may be large and may itself be non-stationary.

The classical Autoregressive Distributed Lag (ARDL) bounds test of Pesaran,
Shin and Smith (2001) works inside a conditional error-correction model

$$\Delta Y_t = a_0 + \rho Y_{t-1} + \theta D_{t-1} + \delta \Delta D_t
  + \sum_i \gamma_i \Delta Y_{t-i} + e_t,$$

with the joint null $H_0: \rho = \theta = 0$ on the lagged levels. Because the
integration order of $D$ is unknown, the test *brackets* it: the lower critical
value treats $D$ as $I(0)$, the upper as $I(1)$, and the statistic is read
against the interval. Outside it the verdict is conclusive; inside it,
inconclusive.

## The second unknown

Now condition on a large control set $W_t$. Projecting the lagged levels onto
controls that themselves carry stochastic trends can **absorb** part of the
long-run variation that identifies the error-correction relation. What then
governs the null is not the integration order of the original regressors but
the number of stochastic trends that survive residualisation. The bracket
reappears, over a different quantity.

```{r bracket, fig.cap = "As the effective integrated count falls, the limiting null slides from the I(1) endpoint to the I(0) endpoint. Classical bounds testing is the right-hand end."}
plot_bracket(k = 10, k_tilde = 6)
```

One consequence is worth stating plainly, because it decides how much of this
you need to care about:

> **Stationary controls are harmless.** A stationary regressor cannot track the
> stochastic trend of an integrated one, so it cannot absorb one. Add a hundred
> and the effective count is unchanged. Only *integrated* controls make the
> bracket live.

## Why no bounds table ships with this package

Tabulated critical values are not valid in this setting, and the
generated-regressor remainder is not negligible at the sample sizes applied work
actually uses. Every critical value here is computed instead: by simulation for
the classical bracket, and by a restricted system wild bootstrap for inference
on real data.

# The data

The package bundles nine monthly United States macroeconomic series from
FRED-MD (McCracken and Ng, 2016), 1973-01 to 2020-12, so every example runs
offline. The application is exchange-rate pass-through: does the dollar pass
through to consumer prices in the long run, conditional on the macro-financial
environment?

```{r data}
df <- passthrough_regime("1999-2007")
dim(df)
CONTROLS
DEFAULT_INTEGRATED
```

Testing within a monetary regime rather than across the whole sample avoids
conflating a structural break with a long-run relationship.

# A first fit

```{r fit}
W <- as.matrix(df[, CONTROLS])
fit <- dml_bounds(df$cpi, df$neer, W,
                  lags = 4, n_blocks = 5, buffer = 6,
                  integrated = DEFAULT_INTEGRATED)
fit
```

Three moving parts sit behind that call.

**A balanced first stage.** The two projections use *different* regressors,
because their targets have different integration orders. $\Delta Y_t$ is
stationary, so the integrated controls enter its projection in first
differences; regressing a stationary target on integrated levels would be
unbalanced and spurious. The tested levels $Z_{t-1} = (Y_{t-1}, D_{t-1})$ are
integrated and are projected on the control *levels*. All trend absorption
happens in that second projection.

```{r design}
des <- build_balanced_design(df$cpi, df$neer, W, lags = 4,
                             integrated = DEFAULT_INTEGRATED)
des
```

**h-block cross-fitting.** The sample is cut into $K$ contiguous chronological
blocks, and each is evaluated by a model trained on the others *minus an
$h$-observation buffer*. The buffer decouples first-stage error from the
evaluation-fold innovations. It costs sample, and the package makes that cost
visible before you commit to a configuration:

```{r sampleuse}
round(sample_use_table(nrow(df)), 3)
```

**A restricted system wild bootstrap.** One Rademacher weight is applied to the
*stacked* pair of conditional and marginal residuals, and both series are
regenerated jointly. This preserves the contemporaneous correlation between
them, which is the endogeneity channel the whole framework exists for. A scheme
that holds $D$ fixed and reweights only the equation error simulates a world
with zero correlation whatever the data say.

```{r boot}
fit <- dml_bootstrap(fit, B = 49, seed = 20260625)
summary(fit)
```

```{r nullplot, fig.cap = "The bootstrap null against the borrowed classical bound. The gap between the two is the argument for not using a table."}
plot(fit)
```

The number of bootstrap draws is kept small here so the vignette builds
quickly. Use `B = 999` for anything you intend to report.

# The diagnostic you must run

The estimand is conditional. If a control is itself part of the equilibrium
system, residualising removes the relation rather than the confounding, and a
non-rejection then means nothing.

`trend_absorption()` runs four fits -- full and reduced control sets crossed
with the adaptive and unpenalised level projection -- and reads the two gaps
together with the stability of the long-run coefficient.

```{r absorb}
ta <- trend_absorption(df$cpi, df$neer, W, drop = REDUCED_DROP,
                       B = 19, seed = 1,
                       lags = 4, n_blocks = 5, buffer = 6,
                       integrated = DEFAULT_INTEGRATED)
ta
```

A verdict that flips when trend-sharing controls are dropped is the signature
of over-absorption, and the reduced-set verdict is then the more credible one.
Concordant verdicts indicate the conclusion is not an artefact. This is a
hypothesis-generating device, not a formal test: it has no size and no power,
and it tells you where to look.

# Penalty sensitivity

A verdict can turn on the penalty and on the projection. Reporting one cell is
specification search; reporting the grid is the method. The column to watch is
`n_selected_Z`, the number of control *levels* retained -- the empirical
counterpart of the effective integrated count.

```{r sweep}
sw <- penalty_sensitivity(df$cpi, df$neer, W,
                          lags_grid = 4, n_blocks = 5, buffer = 6,
                          integrated = DEFAULT_INTEGRATED)
sw
attr(sw, "theta_sign_flips")
```

A long-run coefficient that changes sign across this grid is a warning that the
conditioning set, not the data, is driving the answer.

# Where the critical values come from

The classical bracket is regenerated from the data-generating process Pesaran,
Shin and Smith printed in the notes to their Table CI, rather than transcribed.
That makes it checkable against print, and extends it to any sample size and
past $k = 10$, where the published tables stop.

```{r pss}
sim <- simulate_pss_bounds(k = 1, case = 3, TT = 1000, nsim = 300, seed = 11)
sim
pss_reference(k = 1, case = 3)
```

With only 300 replications the simulated values carry visible Monte Carlo
error; the published table used 40,000. The point of the comparison is that the
generator can be checked, not that 300 draws suffice.

The classical test itself is available for benchmarking, and reads its
statistic against a bracket simulated at *its own* sample size rather than one
calibrated at $T = 1000$:

```{r classical}
cb <- classical_bounds_test(df$cpi, cbind(neer = df$neer), lags = 4,
                            nsim = 300, seed = 11)
cb
```

# Validation

The statistic is deterministic given the specification, so it can be checked
against an independent implementation of the same procedure. On the four
monetary regimes with the full control set, this package and a separate Python
implementation agree to four decimal places:

| regime | this package | reference |
|:---|---:|---:|
| 1973--1985 | 9.111 | 9.111 |
| 1986--1998 | 18.712 | 18.712 |
| 1999--2007 | 0.404 | 0.404 |
| 2008--2020 | 3.703 | 3.703 |

The test suite asserts these values, along with the classical statistics and
the fold structure, so a change that moves any of them fails loudly.

# Limitations

Stated plainly, because they affect how a result should be read.

* **Power is modest at small $T$.** Under integrated nuisance, bootstrap power
  reaches useful levels only from roughly $T = 250$. Much applied ARDL work runs
  at $n = 30\text{--}80$, where this is not the right tool; use the classical
  test with simulated finite-sample bounds instead.
* **Over-absorption is not testable.** The requirement that controls span
  confounding trends but not the cointegrating relation cannot be verified. The
  diagnostic is a signal, not a guarantee.
* **The penalty can change the verdict.** Report the grid, not one cell.
* **The long-run coefficient is less reliable than the verdict.** Across
  regimes it is unstable and often imprecisely estimated, while the test
  decisions are stable.
* **The integrated block must stay small.** The validity theory keeps it
  fixed-dimensional; selecting among many integrated regressors is an open
  problem, and the package warns when you cross a threshold.

# References

Chernozhukov, V., Chetverikov, D., Demirer, M., Duflo, E., Hansen, C., Newey,
W. and Robins, J. (2018). Double/debiased machine learning for treatment and
structural parameters. *The Econometrics Journal*, 21(1), C1--C68.
doi:10.1111/ectj.12097

McCracken, M. W. and Ng, S. (2016). FRED-MD: A monthly database for
macroeconomic research. *Journal of Business & Economic Statistics*, 34(4),
574--589. doi:10.1080/07350015.2015.1086655

Narayan, P. K. (2004). Reformulating critical values for the bounds
F-statistics approach to cointegration. Monash University Discussion Paper
02/04.

Pesaran, M. H., Shin, Y. and Smith, R. J. (2001). Bounds testing approaches to
the analysis of level relationships. *Journal of Applied Econometrics*, 16(3),
289--326. doi:10.1002/jae.616

Villena, M. J. (2026). Testing cointegration with many persistent controls.
SSRN working paper. doi:10.2139/ssrn.6472826

Zou, H. (2006). The adaptive lasso and its oracle properties. *Journal of the
American Statistical Association*, 101(476), 1418--1429.
doi:10.1198/016214506000000735
